---
title: "ggplot2 examples and exercises part 2"
output: html_document
---

```{r setup, include=FALSE, warning=FALSE, message=FALSE}
knitr::opts_chunk$set(echo = TRUE, cache = TRUE)
library(ggplot2)

```


## Extensions
While ggplot2 comes with a lot of batteries included, the extension ecosystem 
provides priceless additional features

### Plot composition
We start by creating 3 separate plots
```{r}
p1 <- ggplot(msleep) + 
  geom_boxplot(aes(x = sleep_total, y = vore, fill = vore))

p1
```

```{r}
p2 <- ggplot(msleep) + 
  geom_bar(aes(y = vore, fill = vore))

p2
```

```{r}
p3 <- ggplot(msleep) + 
  geom_point(aes(x = bodywt, y = sleep_total, colour = vore)) + 
  scale_x_log10()

p3
```

Combining them with patchwork is a breeze using the different operators

```{r}
library(patchwork)

p1 + p2 + p3

(p1 | p2) / 
   p3

p_all <- (p1 | p2) / 
            p3
p_all + plot_layout(guides = 'collect')

p_all & theme(legend.position = 'none')    # the & operator applies this to all plots, unlike the + operator

p_all <- p_all & theme(legend.position = 'none')
p_all + plot_annotation(
  title = 'Mammalian sleep patterns',
  tag_levels = 'A'
)
```

#### Exercises
Patchwork will assign the same amount of space to each plot by default, but this
can be controlled with the `widths` and `heights` argument in `plot_layout()`. 
This can take a numeric vector giving their relative sizes (e.g. `c(2, 1)` will 
make the first plot twice as big as the second). Modify the code below so that
the middle plot takes up half of the total space:

```{r}
p <- ggplot(mtcars) + 
  geom_point(aes(x = disp, y = mpg))
p + p + p
```

```{r}
p + p + p + plot_layout(widths = c(1, 2, 1))
```

```{r}
# Another extension - using actual units for the width:
p + p + p + plot_layout(widths = unit(c(5, 1, 1), c('cm', 'null', 'null')))  # first plot gets a fixed width of 5cm, the two remaining plots then split the remaining space between them
```


* * *

The `&` operator can be used with any type of ggplot2 object, not just themes.
Modify the code below so the two plots share the same y-axis (same limits)

```{r}
p1 <- ggplot(mtcars[mtcars$gear == 3,]) + 
  geom_point(aes(x = disp, y = mpg))
p2 <- ggplot(mtcars[mtcars$gear == 4,]) + 
  geom_point(aes(x = disp, y = mpg))

p1 + p2
```

```{r}
p1 <- ggplot(mtcars[mtcars$gear == 3,]) + 
  geom_point(aes(x = disp, y = mpg))
p2 <- ggplot(mtcars[mtcars$gear == 4,]) + 
  geom_point(aes(x = disp, y = mpg))

p1 + p2 & scale_y_continuous(limits = c(10, 40))
```


* * *

Patchwork contains many features for fine tuning the layout and annotation. Very
complex layouts can be obtained by providing a design specification to the 
`design` argument in `plot_layout()`. The design can be defined as a textual 
representation of the cells. Use the layout given below. How should the textual 
representation be understood.

```{r}
p1 <- ggplot(mtcars) + 
  geom_point(aes(x = disp, y = mpg))
p2 <- ggplot(mtcars) + 
  geom_bar(aes(x = factor(gear)))
p3 <- ggplot(mtcars) + 
  geom_boxplot(aes(x = factor(gear), y = mpg))

layout <- '
AA#
#BB
C##
'
# Note that 1,2,3 would also work instead of A, B, C here

p1 + p2 + p3 + plot_layout(design = layout)
```

### Animation
ggplot2 is usually focused on static plots, but gganimate extends the API and
grammar to describe animations. As such it feels like a very natural extension
of using ggplot2

```{r}
ggplot(economics) + 
  geom_line(aes(x = date, y = unemploy))

library(gganimate)

ggplot(economics) + 
  geom_line(aes(x = date, y = unemploy)) + 
  transition_reveal(along = date)
```

There are many different transitions that control how data is interpreted for
animation, as well as a range of other animation specific features

```{r}
ggplot(mpg) + 
  geom_bar(aes(x = factor(cyl)))

ggplot(mpg) + 
  geom_bar(aes(x = factor(cyl))) + 
  labs(title = 'Number of cars in {closest_state} by number of cylinders') + # gives a dynamic chart title
  transition_states(states = year) + 
  enter_grow() + # when new data comes into the animation, we will use grow
  exit_fade()  # when data leaves the animation we will just fade it out
```

#### Exercises
The animation below will animate between points showing cars with different 
cylinders.

```{r}
ggplot(mpg) + 
  geom_point(aes(x = displ, y = hwy)) + 
  ggtitle("Cars with {closest_state} cylinders") + 
  transition_states(factor(cyl))
```

gganimate uses the `group` aesthetic to match observations between states. By 
default the group aesthetic is set to the same value, so observations are 
matched by their position (first row of 4 cyl is matched to first row of 5 cyl 
etc.). This is clearly wrong here (why?). Add a mapping to the `group` aesthetic
to ensure that points do not move between the different states.


```{r}
ggplot(mpg) + 
  geom_point(aes(x = displ, y = hwy, group = seq_len(nrow(mpg)))) + 
  ggtitle("Cars with {closest_state} cylinders") + 
  transition_states(factor(cyl))
```


* * *

In the presence of discrete aesthetic mappings (`colour` below), the group is
deduced if not given. The default behaviour of objects that appear and disappear
during the animation is to simply pop in and out of existance. `enter_*()` and
`exit_*()` functions can be used to control this behaviour. Experiment with the
different enter and exit functions provided by gganimate below. What happens if 
you add multiple enter or exit functions to the same animation?

```{r}
ggplot(mpg) + 
  geom_point(aes(x = displ, y = hwy, colour = factor(cyl))) + 
  ggtitle("Cars with {closest_state} cylinders") + 
  transition_states(factor(cyl))
```

* * *

In the animation below (as in all the other animations) the changes happens at
constant speed. How values change during an animation is called easing and can
be controlled using the `ease_aes()` function. Read the documentation for 
`ease_aes()` and experiment with different easings in the animation.

```{r}
mpg2 <- tidyr::pivot_longer(mpg, c(cty,hwy))

ggplot(mpg2) + 
  geom_point(aes(x = displ, y = value)) + 
  ggtitle("{if (closest_state == 'cty') 'Efficiency in city' else 'Efficiency on highway'}") +
  transition_states(name)
```

```{r}
mpg2 <- tidyr::pivot_longer(mpg, c(cty,hwy))

ggplot(mpg2) + 
  geom_point(aes(x = displ, y = value)) + 
  ggtitle("{if (closest_state == 'cty') 'Efficiency in city' else 'Efficiency on highway'}") +
  transition_states(name) +
  ease_aes(y = 'cubic-in-out')  # the graph movement is no longer linear, it speeds up and down...this is nicer to look at but beware not to use this when moving along a time axis!
```


### Annotation
Text is a huge part of storytelling with your visualisation. Historically, 
textual annotations has not been the best part of ggplot2 but new extensions 
make up for that.

Standard geom_text will often result in overlaping labels
```{r}
ggplot(mtcars, aes(x = disp, y = mpg)) + 
  geom_point() + 
  geom_text(aes(label = row.names(mtcars)))
```

ggrepel takes care of that

```{r}
library(ggrepel)
ggplot(mtcars, aes(x = disp, y = mpg)) + 
  geom_point() + 
  geom_text_repel(aes(label = row.names(mtcars)))
```

If you want to highlight certain parts of your data and describe it, the 
`geom_mark_*()` family of geoms have your back

```{r}
library(ggforce)
ggplot(mtcars, aes(x = disp, y = mpg)) +
  geom_point() + 
  geom_mark_ellipse(aes(filter = gear == 4,
                        label = '4 gear cars',
                        description = 'Cars with fewer gears tend to both have higher yield and lower displacement'))
```

#### Exercises
ggrepel has a tonne of settings for controlling how text labels move. Often, 
though, the most effective is simply to not label everything. There are two
strategies for that: Either only use a subset of the data for the repel layer,
or setting the label to `""` for those you don't want to plot. Try both in the 
plot below where you only label 10 random points.

```{r}
mtcars2 <- mtcars
mtcars2$label <- rownames(mtcars2)
points_to_label <- sample(nrow(mtcars), 10)
ggplot(mtcars2, aes(x = disp, y = mpg)) + 
  geom_point() + 
  geom_text_repel(aes(label = label))
```

```{r}
mtcars2 <- mtcars
mtcars2$label <- rownames(mtcars2)
points_to_label <- sample(nrow(mtcars), 10)

ggplot(mtcars2, aes(x = disp, y = mpg)) + 
  geom_point() + 
  geom_text_repel(aes(label = label), data = mtcars2[points_to_label,])
```

```{r}

mtcars2 <- mtcars
mtcars2$label <- rownames(mtcars2)
points_to_label <- sample(nrow(mtcars), 10)


mtcars2$label[-points_to_label] <- ""  # Stop overlapping of labels with points
ggplot(mtcars2, aes(x = disp, y = mpg)) + 
  geom_point() + 
  geom_text_repel(aes(label = label))



```


* * *

Explore the documentation for `geom_text_repel`. Find a way to ensure that the 
labels in the plot below only repels in the vertical direction

```{r}
mtcars2$label <- ""
mtcars2$label[1:10] <- rownames(mtcars2)[1:10]
ggplot(mtcars2, aes(x = disp, y = mpg)) + 
  geom_point() + 
  geom_text_repel(aes(label = label))
```

```{r}
mtcars2$label <- ""
mtcars2$label[1:10] <- rownames(mtcars2)[1:10]
ggplot(mtcars2, aes(x = disp, y = mpg)) + 
  geom_point() + 
  geom_text_repel(aes(label = label), direction = 'y')
```


* * *

ggforce comes with 4 different types of mark geoms. Try them all out in the code
below:

```{r}
ggplot(mtcars, aes(x = disp, y = mpg)) +
  geom_point() + 
  geom_mark_ellipse(aes(filter = gear == 4,
                        label = '4 gear cars'))

# geom_mark_rect
ggplot(mtcars, aes(x = disp, y = mpg)) +
  geom_point() + 
  geom_mark_rect(aes(filter = gear == 4,
                        label = '4 gear cars'))

ggplot(mtcars, aes(x = disp, y = mpg)) +
  geom_point() + 
  geom_mark_rect(aes(filter = gear == 4,
                        label = '4 gear cars'), expand = 0)

ggplot(mtcars, aes(x = disp, y = mpg)) +
  geom_point() + 
  geom_mark_rect(aes(filter = gear == 4,
                        label = '4 gear cars'), radius = 0)

#geom_mark_circle
ggplot(mtcars, aes(x = disp, y = mpg)) +
  geom_point() + 
  geom_mark_circle(aes(filter = gear == 4,
                        label = '4 gear cars'))

ggplot(mtcars, aes(x = disp, y = mpg)) +
  geom_point() + 
  geom_mark_circle(aes(filter = gear == 4,
                        label = '4 gear cars')) +
  coord_cartesian(clip = 'off')


# geom_mark_hull
ggplot(mtcars, aes(x = disp, y = mpg)) +
  geom_point() + 
  geom_mark_hull(aes(filter = gear == 4,
                        label = '4 gear cars'), concavity = 7)

ggplot(mtcars, aes(x = disp, y = mpg)) +
  geom_point() +
  geom_mark_hull(aes(fill = factor(gear), label = gear))

ggplot(mtcars, aes(x = disp, y = mpg)) +
  geom_point() +
  geom_mark_hull(aes(group = factor(gear), label = gear), fill = NA, colour = NA)
```

### Networks
ggplot2 has been focused on tabular data. Network data in any shape and form is
handled by ggraph

```{r}
library(ggraph)  # Thomoas Pedersen's packages, gives best capabilities
library(tidygraph)  # These 2 packages usually come together

graph <- create_notable('zachary') %>% 
  mutate(clique = as.factor(group_infomap()))

ggraph(graph) + 
  geom_edge_link() + 
  geom_node_point(aes(colour = clique), size = 2)

ggraph(graph) + 
  geom_mark_hull(aes(x, y, fill = clique)) + 
  geom_edge_link() + 
  geom_node_point(aes(colour = clique), size = 2)

```

dendrograms are just a specific type of network

```{r}
iris_clust <- hclust(dist(iris[, 1:4]))

ggraph(iris_clust) + 
  geom_edge_bend() + 
  geom_node_point(aes(filter = leaf))
```

#### Exercises
Most network plots are defined by a layout algorithm, which takes the network 
structure and calculate a position for each node. The layout algorithm is 
global and set in the `ggraph()`. The default `auto` layout will inspect the
network object and try to choose a sensible layout for it (e.g. dendrogram for a
hierarchical clustering as above). There is, however no optimal layout and it is
often a good idea to try out different layouts. Try out different layouts in the
graph below. See the [the website](https://ggraph.data-imaginist.com/reference/index.html)
for an overview of the different layouts.

```{r}
ggraph(graph) + 
  geom_edge_link() + 
  geom_node_point(aes(colour = clique), size = 3)
```

```{r}
ggraph(graph, layout = 'gem') + 
  geom_edge_link() + 
  geom_node_point(aes(colour = clique), size = 3)
```


* * *

There are many different ways to draw edges. Try to use `geom_edge_parallel()` 
in the graph below to show the presence of multiple edges

```{r}
highschool_gr <- as_tbl_graph(highschool)
ggraph(highschool_gr) + 
  geom_edge_link() + 
  geom_node_point()

#highschool_gr <- as_tbl_graph(highschool)
ggraph(highschool_gr) + 
  geom_edge_parallel() + 
  geom_node_point()

ggraph(highschool_gr) + 
  geom_edge_fan() + 
  geom_node_point()
```

Faceting works in ggraph as it does in ggplot2, but you must choose to facet by
either nodes or edges. Modify the graph below to facet the edges by the `year` 
variable (using `facet_edges()`)

```{r}
ggraph(highschool_gr) + 
  geom_edge_fan() + 
  geom_node_point()

ggraph(highschool_gr) + 
  geom_edge_fan() + 
  geom_node_point() +
  facet_edges(~year)

```

### Looks
Many people have already desgned beautiful (and horrible) themes for you. Use
them as a base

```{r}
p <- ggplot(mtcars, aes(mpg, wt)) +
  geom_point(aes(color = factor(carb))) +
  labs(
    x = 'Fuel efficiency (mpg)', 
    y = 'Weight (tons)',
    title = 'Seminal ggplot2 example',
    subtitle = 'A plot to show off different themes',
    caption = 'Source: It’s mtcars — everyone uses it'
  )

library(hrbrthemes)  # great starting point for nice themes

p + 
  scale_colour_ipsum() + 
  theme_ipsum()
```

```{r}
library(ggthemes)
p + 
  scale_colour_excel() + 
  theme_excel()  # Looks like an Excel graph
```

## Drawing anything

```{r}
states <- c(
  'eaten', "eaten but said you didn\'t", 'cat took it', 'for tonight',
  'will decompose slowly'
)

pie <- data.frame(
  state = factor(states, levels = states),
  amount = c(4, 3, 1, 1.5, 6),
  stringsAsFactors = FALSE
)

ggplot(pie) + 
  geom_col(aes(x = 0, y = amount, fill = state))
```

```{r}
ggplot(pie) + 
  geom_col(aes(x = 0, y = amount, fill = state)) + 
  coord_polar(theta = 'y')
```

```{r}
ggplot(pie) + 
  geom_col(aes(x = 0, y = amount, fill = state)) + 
  coord_polar(theta = 'y') + 
  scale_fill_tableau(name = NULL,
                     guide = guide_legend(ncol = 2)) + 
  theme_void() + 
  theme(legend.position = 'top', 
        legend.justification = 'left')
```

```{r}
ggplot(pie) + 
  geom_arc_bar(aes(x0 = 0, y0 = 0, r0 = 0, r = 1, amount = amount, fill = state), stat = 'pie') + 
  coord_fixed()
```

```{r}
ggplot(pie) + 
  geom_arc_bar(aes(x0 = 0, y0 = 0, r0 = 0, r = 1, amount = amount, fill = state), stat = 'pie') + 
  coord_fixed() + 
  scale_fill_tableau(name = NULL,
                     guide = guide_legend(ncol = 2)) + 
  theme_void() + 
  theme(legend.position = 'top', 
        legend.justification = 'left')
```

```{r}
ggplot(mpg) + 
  # geom_bar(aes(x = hwy), stat = 'bin')
  geom_histogram(aes(x = hwy))

ggplot(mpg) + 
  geom_bar(aes(x = hwy), stat = 'bin')
#  geom_histogram(aes(x = hwy))
```

```{r}
ggplot(mpg) + 
  geom_bar(aes(x = hwy)) + 
  scale_x_binned(n.breaks = 30, guide = guide_axis(n.dodge = 2))
```